---
title: "DNA contamination QC"
output:
  html_document:
    theme: yeti # A clean and modern theme from Bootswatch
    highlight: tango  # Syntax highlighting style
    code_folding: hide  # Collapsible code blocks
    toc: true  # Include a table of contents
    toc_depth: 3  # Show TOC up to level 3 headings
    toc_float: true  # Make the TOC float while scrolling
    css: custom.css  # Link to your custom CSS file
params:
  args: ''
  dark_mode: true  # Add parameter for enabling dark mode
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(plotly)

color_palette <- c(
  "#E69F00", # Orange
  "#56B4E9", # Sky Blue
  "#009E73", # Green
  "#F0E442", # Yellow
  "#0072B2", # Blue
  "#D55E00", # Vermilion
  "#CC79A7", # Reddish Purple
  "#999999", # Gray
  "#000000", # Black
  "#A6761D", # Brown
  "#1B9E77", # Dark Green
  "#D95F02", # Dark Orange
  "#7570B3", # Violet
  "#E7298A", # Magenta
  "#66A61E"  # Lime Green
)
args <-  params$args
output_dir <- args[1]

samples <- list.files(output_dir)
annotation <- read_table("../../resources/annotation.tsv")

```

# Quality control
## DNA contamination
On of the largest caveats in RNA sequencing, and extracellular RNA sequencing especially, is DNA contamination. Luckily, we can determine the extent of this problem by investigation the metrics of our output. None of these metrics on their own can determine DNA contamination but, together, they can indicate an issue.

### Strandedness
Many RNA sequencing library preparation protocols use a stranded approach. This means that the reads generated from a transcript map specifically to either the coding or template strand. The SMARTer® Stranded Total RNA-Seq Kit - Pico Input Mammalian, for example, is a stranded protocol. More specifically, it is a reverse stranded protocol, meaning that the reads map to the template strand. When sequencing DNA, both strands will be sequenced and your stranded library preparation becomes unstranded. Ideally, the fraction of "reads mapped to template strand" is close to 1. The closer this value gets to 0.5, the more likely your data contains some DNA data.

```{r strandedness, echo = FALSE}
strandedness <- tibble()

for (sample in samples) {
  x <- read_table(paste0(output_dir, "/", sample, "/08_rseqc/", sample, "_strandedness.txt"), skip = 3, col_names = FALSE)$X7
  strandedness <- rbind(strandedness, 
                        tibble(sample = sample, 
                               type = factor(c("unsure", "mapped to coding strand", "mapped to template strand"), levels = c("unsure", "mapped to template strand", "mapped to coding strand")), 
                               fraction = x)
  )
}

ggplotly(strandedness %>%
  ggplot(aes(sample, fraction, fill = type)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  theme(axis.title = element_text(face = "bold"), 
        legend.title = element_blank(),
        axis.text.x = element_text(angle = 45)) +
  xlab("") + 
  ylab("fraction of\ndeduplicated reads\n") +
  scale_fill_manual(values = color_palette))
```

### Splice reads
Since the kits used here are total RNA sequencing kits, we at least expect some of the reads to map to splice junctions. In case of DNA, there should be little to no reads mapping to splice junctions. 

```{r spliced_reads, echo = FALSE}
spliced_reads <- tibble()

for (sample in samples) {
  x <- read_table(paste0(output_dir, "/", sample, "/08_rseqc/", sample, "_raw_bam_stat.txt"), col_names = paste0("X", 1:5))
  spliced_reads <- rbind(spliced_reads,
                         tibble(type = c("non-splice", "splice"), 
                                reads = as.double(x$X3[15:16]), 
                                sample = sample)
                         )
}

ggplotly(spliced_reads %>%
  group_by(sample) %>%
  mutate(fraction = reads / sum(reads)) %>%
  ungroup() %>%
  ggplot(aes(sample, fraction, fill = type)) +
  geom_bar(stat = "identity") +
  theme_minimal() + 
  theme(axis.title = element_text(face = "bold"), 
        legend.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("") + 
  ylab("fraction of raw reads\n") +
  scale_fill_manual(values = color_palette))
```

## Biotype distribution
```{r biotype_distribution, echo = FALSE}
counts <- tibble()
for (sample in samples) {
  x <- read_table(paste0(output_dir, "/", sample, "/05_count/", sample, "_htseq_dedup_counts.txt"), col_names = FALSE)
  counts <- rbind(counts,
  x %>% mutate(sample = sample))
}

ggplotly(counts %>%
  filter(grepl("ENSG", X1)) %>%
  left_join(annotation, by = c("X1" = "gene_id")) %>%
  group_by(sample, biotype) %>%
  summarise(count = sum(X2)) %>%
  ungroup() %>%
  group_by(sample) %>%
  mutate(fr = count / sum(count)) %>%
  ungroup() %>%
  mutate(new_biotype = ifelse(fr < 0.005, "less than 1%", biotype)) %>%
  group_by(new_biotype) %>%
  mutate(avg_over_biotype = mean(fr)) %>%
  ungroup() %>%
  arrange(desc(avg_over_biotype)) %>%
  mutate(new_biotype = factor(new_biotype, levels = unique(new_biotype))) %>% 
  group_by(sample, new_biotype) %>%
  summarise(fr = sum(fr)) %>%
  ungroup() %>%
  ggplot(aes(sample, fr, fill = new_biotype)) +
  geom_histogram(stat = "identity") +
  theme_minimal() + 
  theme(axis.title = element_text(face = "bold"), 
        legend.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("") + 
  ylab("fraction of counts\n") +
  scale_fill_manual(values = color_palette))
```

## Count distribution
```{r counts_distribution, echo = FALSE}
ggplotly(counts %>%
  mutate(type = ifelse(grepl("__", X1), X1, 
                  ifelse(grepl("^R", X1), "sequin",
                    ifelse(grepl("^ERCC", X1), "ERCC", "genetic")))) %>%
  mutate(type = factor(type, levels = c("__no_feature", "__ambiguous", "__too_low_aQual", "__not_aligned", "__alignment_not_unique", "sequin", "ERCC", "genetic"))) %>%
  group_by(sample, type) %>%
  summarise(count = sum(X2)) %>%
  ungroup() %>%
  group_by(sample) %>%
  mutate(fr = count / sum(count)) %>%
  ungroup() %>%
  ggplot(aes(sample, fr, fill = type)) + 
  geom_histogram(stat = "identity") +
  theme_minimal() + 
  theme(axis.title = element_text(face = "bold"), 
        legend.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("") + 
  ylab("fraction of counts\n") +
  scale_fill_manual(values = color_palette))

ggplotly(counts %>%
  filter(!grepl("__|^R|ERCC", X1)) %>%
  dplyr::rename(`log10(counts)` = X2) %>%
  mutate(counts = 10^`log10(counts)`) %>%
  ggplot(aes(`log10(counts)`, col = sample)) +
  stat_ecdf(aes(`log10(counts)`, col = sample)) +
  theme_minimal() +
  theme(axis.title = element_text(face = "bold"),
        legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("\ncounts") +
  ylab("fraction of genes\n") +
  scale_color_manual(values = rep(color_palette, 50)) +
  scale_x_log10())
```

